R-LIME: LIME法における矩形制約と最適化

### 2024-03-19 paper/paper.tex
ICPR 4th version

\ifnum\mode=0
  \pagestyle{plain}
  \pagenumbering{roman}
  \tableofcontents
  \newpage
  \listoffigures\listoftables\clearpage
  \AddToHook{cmd/section/before}{\clearpage}
  \pagestyle{plain}
  \pagenumbering{arabic}
\else
  % The abstract should briefly summarize the contents of the paper in
  % 150--250 words.
  \begin{abstract}
    In recent years,
    complex machine learning models such as deep neural networks
    have been used in various industrial fields due to their high accuracy.
    However,
    its complexity has been a major obstacle to implementation
    in decision-making situations where transparency of the decision process
    is required.
    In order to address this problem,
    various post-hoc explanation methods have been proposed,
    but they have not been able to achieve interpretability of
    both the explanation and its scope.
    We propose a new method, R-LIME,
    which interprets complex classifiers in an interpretable scope.
    R-LIME locally approximizes a complex decision boundary linearly
    in a rectangular region
    and maximizes the region as long as the accuracy of the linear
    classifier is higher than a specified threshold.
    The resulting rectangular region is interpretable for users because it is
    expressed as a conjunction of feature predicates.
    We demonstrate the effectiveness of the proposed method through qualitative
    and quantitative experiments using a real-world dataset.
    Finally, we discuss limitations and future work of the proposed method.
    % 148 words
    \keywords{Interpretable machine learning \and Local surrogate model}
  \end{abstract}
\fi

@@

Then, model-agnostic methods are classified into
\emph{global} and \emph{local} methods based on their locality in input space.
Global methods,
such as Partial Dependence Plots (PDP)~\cite{friedman2001greedy}
and Accumulated Local Effects (ALE)~\cite{apley2020visualizing},
aim to explain the model's behavior across the entire input space.
% seeking explanations valid for any input.
However, providing global explanations becomes challenging
as the model's complexity increases.
In contrast, local methods,
such as Local Interpretable Model-agnostic Explanations (LIME)
\cite{ribeiro2016why}
and Anchor~\cite{ribeiro2018anchors},
% SHAP~\cite{lundberg2017unified},
% EXPLAN~\cite{rasouli2020explan},
% LORE (Local Rule-based Explanations)~\cite{guidotti2018local}
% and BELLA (Black box model Explanations by Local Linear Approximations)
% \cite{radulovic2023bella},
explain model's behavior in the vicinity of a specific input.
While local methods offer explanations more simple and accurate
than global methods,
the scope of the explanation is limited locally.

@@

\ifnum\mode=1
  \subsection{Computational Complexity}
  % \subsubsection{Number of Queries}
  Post-hoc explanation methods including LIME, Anchor, and R-LIME
  need to query the black-box model multiple times
  to get labels for perturbed samples, which is computationally expensive.
  The number of queries required for LIME is $|\mathcal{Z}_p|$,
  where $|\mathcal{Z}_p|$ is the number of perturbed samples selected by the user.
  On the other hand,
  the number of queries required for Anchor and R-LIME is
  $\mathcal{O}(B\cdot m^2+m^2\cdot\mathcal{O}_{\mathrm{MAB}[B\cdot m,B]})$,
  where $B$ is the beam width, $m$ is the number of features, and
  $\mathcal{O}_{\mathrm{MAB}[B\cdot m,B]}$ is the computational complexity of
  best arm identification finding the $B$ best arms among $B\cdot m$ arms.
  $\mathcal{O}_{\mathrm{MAB}[B\cdot m,B]}$ depends on hyperparameters
  such as beam width $B$, confidence level $1-\delta$ and tolerance $\epsilon$.
  % \subsubsection{Updating the Linear Classifier}
  % For Anchor, each trial is composed of sampling a perturbation vector $z$
  % and getting output $f(z)$.
  % But for R-LIME, a trial
  % includes additionally updating the linear classifier $g$ by $z$ and $f(z)$.
  % If we use support vector machine (SVM) as the linear classifier and
  % update it by stochastic gradient descent (SGD)~\cite{robbins1951stochastic},
  % the computational complexity of updating $g$ for each trial is $\mathcal{O}(m^2)$.
\fi

@@

\subsection{Quantitative Evaluation}\label{sec:exp2}
\subsubsection{Experimental Setup}
To demonstrate that R-LIME learns a highly accurate linear approximation model
in the optimized approximation region,
we conducted a comparison of the local accuracy of explanations
between LIME and R-LIME\@.
Using the same settings as in \cref{sec:exp_setting},
we randomly sampled 100 instances from the test data of the recidivism dataset
and generated explanations (with $\tau=0.70,0.80,0.90$) using LIME and R-LIME\@.
We then sampled 10,000 instances within the rectangular region
obtained by R-LIME and calculated the local accuracy of both methods.
\subsubsection{Experimental Results}
\begin{figure}[tbp]
  \centering
  \includegraphics[width=0.53\textwidth]{experiments/exp2/box_plot}
  \caption[Comparison of Local Accuracy between R-LIME and LIME]{%
    Comparison of local accuracy between LIME and R-LIME\@.
    R-LIME achieved higher and less variable accuracy compared to LIME
    for all values of $\tau$.
  }\label{fig:box_plot}
\end{figure}
The results are presented in \cref{fig:box_plot},
showing the distribution of the accuracy of the linear approximation models
learned by LIME and R-LIME\@.
R-LIME achieved higher accuracy compared to LIME for all values of $\tau$.
This suggests that the linear classifiers learned by LIME and R-LIME
differ significantly,
and R-LIME learns a high-accuracy linear classifier adapted to the rectangular region.
Additionally, as $\tau$ increases,
the variability in the accuracy of LIME widens.
This indicates that the linear classifiers learned by LIME may not function
effectively as approximation models depending on how the region is selected.

\section{Challenges and Future Perspectives}
 {%
  \def\imgwidth{0.47\textwidth}
  \begin{figure}[t]
    \centering
    \begin{subfigure}[t]{\imgwidth}
      \centering
      \includegraphics[width=0.75\textwidth]{visual-rlime3}
      \caption{R-LIME for balanced label distribution.}\label{fig:balanced}
    \end{subfigure}
    \begin{subfigure}[t]{\imgwidth}
      \centering
      \includegraphics[width=0.75\textwidth]{visual-rlime-imbalanced}
      \caption{R-LIME for imbalanced label distribution.
      }\label{fig:imbalanced}
    \end{subfigure}
    \caption[Behavior of R-LIME for balanced and imbalanced label distribution]{%
      Behavior of R-LIME for balanced and imbalanced label distribution.
      In case of imbalanced label distribution,
      the approximation region covers the entire input space and the
      linear approximation model always outputs the majority label.
    }
  \end{figure}
 }

@@

\section{Conclusion}
We identified challenges in existing methods for local model-agnostic post-hoc
explanations of black-box classifiers and proposed R-LIME to address them.
We represented the rectangular region for local linear approximation as a
conjunction of feature predicates and proposed an algorithm to
maximize coverage under the constraint of minimum approximation accuracy.
Comparing the outputs of LIME and R-LIME on the real-world dataset,
we demonstrated that R-LIME provides a clear application scope of the
explanation and can be evaluated by users for its reliability and generality.
However, we discussed the instability of behavior concerning imbalanced label
distributions and raised questions about the theoretical validity of using
the KL-LUCB algorithm.

>>>>

\ifnum\mode=0
  \pagestyle{plain}
  \pagenumbering{roman}
  \tableofcontents
  \newpage
  \listoffigures\listoftables\clearpage
  \AddToHook{cmd/section/before}{\clearpage}
  \pagestyle{plain}
  \pagenumbering{arabic}
\else
  % The abstract should briefly summarize the contents of the paper in
  % 150--250 words.
  \begin{abstract}
    In recent years,
    complex machine learning models have been introduced
    in various industrial fields due to their high accuracy.
    However,
    their complexity has been a major obstacle to implementation
    in sensitive decision-making situations.
    In order to address this problem,
    various post-hoc explanation methods have been proposed,
    but they have not been able to achieve interpretability of
    both the explanation and its scope.
    We propose a new method, R-LIME,
    which interprets complex classifiers in an interpretable scope.
    R-LIME locally approximates a decision boundary of a black-box classifier
    in a rectangular region
    and maximizes the region as long as the accuracy of the linear
    classifier is higher than a given threshold.
    The resulting rectangular region is interpretable for users because it is
    expressed as a conjunction of feature predicates.
    Comparing R-LIME with the existing method,
    we demonstrate the effectiveness of the proposed method through qualitative
    and quantitative experiments using a real-world dataset.
    Finally, we discuss limitations and future work of the proposed method.
    % 161 words
    \keywords{Interpretable machine learning \and Local surrogate model}
  \end{abstract}
\fi

@@

Furthermore, model-agnostic methods are categorized into
\emph{global} and \emph{local} methods based on their locality in input space.
Global methods,
such as Partial Dependence Plots (PDP)~\cite{friedman2001greedy}
and Accumulated Local Effects (ALE)~\cite{apley2020visualizing},
aim to explain the model's behavior across the entire input space.
% seeking explanations valid for any input.
However, providing global explanations becomes challenging
as the model's complexity increases.
In contrast, local methods,
such as Local Interpretable Model-agnostic Explanations (LIME)
\cite{ribeiro2016why}
and Anchor~\cite{ribeiro2018anchors},
% SHAP~\cite{lundberg2017unified},
% EXPLAN~\cite{rasouli2020explan},
% LORE (Local Rule-based Explanations)~\cite{guidotti2018local}
% and BELLA (Black box model Explanations by Local Linear Approximations)
% \cite{radulovic2023bella},
explain model's behavior in the vicinity of a specific input.
While local methods offer explanations more simple and accurate
than global methods,
the scope of the explanation is limited locally.

@@

\ifnum\mode=1
  \subsection{Computational Complexity}
  % \subsubsection{Sample Complexity}
  Post-hoc explanation methods including LIME, Anchor, and R-LIME
  need to sample a perturbation vector
  and get the output of the black-box model multiple times,
  which is computationally expensive.
  The number of samples required for LIME is $|\mathcal{Z}_p|$,
  which is the number of samples designated by the user.
  On the other hand,
  the expected number of samples required for Anchor and R-LIME is bounded by
  $\mathcal{O}[m\cdot\mathcal{O}_{\mathrm{MAB}[B\cdot m,B]}]$,
  where $\mathcal{O}_{\mathrm{MAB}[B\cdot m,B]}$
  is the expected number of samples
  for best arm identification finding the best $B$ arms from $B\cdot m$ arms.
  For KL-LUCB algorithm~\cite{kaufmann2013information},
  the expected number of samples is bounded by
  % \begin{equation}
  %   \mathcal{O}\left[H_{\epsilon}\log\frac{H_\epsilon}{\delta}\right],
  % \end{equation}
  % where $H_{\epsilon}$ is sample complexity of the problem
  % \cite{kaufmann2013information}.
  \begin{equation}
    \mathcal{O}_{\mathrm{MAB}[B\cdot m,B]}=
    \mathcal{O}\left[\frac{Bm}{\epsilon^2}\log\frac{Bm}{\epsilon^2\delta}\right].
  \end{equation}

  % \subsubsection{Complexity of Each Trial}
  Each trial of KL-LUCB algorithm consists of sampling a perturbation vector $z$,
  getting output $f(z)$, and updating the linear classifier $g$.
  If we sample a perturbation vector from a multivariate normal distribution
  using Cholesky decomposition in advance ($\mathcal{O}(2^m\cdot m^3)$),
  we can sample $z$ in $\mathcal{O}(m^2)$.
  If we use random forest as the black-box model,
  the complexity of getting output is $\mathcal{O}(1)$.
  If we use logistic regression as the linear classifier and
  update it by stochastic gradient descent (SGD)~\cite{robbins1951stochastic},
  the computational complexity of updating $g$ for each trial is
  $\mathcal{O}(m)$.
  Then the computational complexity of each trial is $\mathcal{O}(m^2)$.

  % \subsubsection{Total Complexity}
  Overall, the complexity of R-LIME is bounded by
  \begin{equation}
    \mathcal{O}\left[
      2^m\cdot m^3+\frac{Bm^4}{\epsilon^2}\log\frac{Bm}{\epsilon^2\delta}
      \right],
  \end{equation}
  which is very sensitive to the number of features $m$.
\fi

@@

\subsection{Quantitative Evaluation}\label{sec:exp2}
\subsubsection{Experimental Setup}
To demonstrate that R-LIME learns a highly accurate linear approximation model
in the optimized approximation region,
we conducted a comparison of the local accuracy of explanations
between LIME and R-LIME\@.
Using the same settings as in \cref{sec:exp_setting},
we randomly sampled 100 instances from the test data of the recidivism dataset
and generated explanations (with $\tau=0.70,0.80,0.90$) using LIME and R-LIME\@.
We then sampled 10,000 instances within the rectangular region
obtained by R-LIME and calculated the local accuracy of both methods.
\subsubsection{Experimental Results}
\begin{figure}[tbp]
  \centering
  \includegraphics[width=0.5\textwidth]{experiments/exp2/box_plot}
  \caption[Comparison of Local Accuracy between R-LIME and LIME]{%
    Comparison of local accuracy between LIME and R-LIME\@.
    R-LIME achieved higher and less variable accuracy compared to LIME
    for all values of $\tau$.
  }\label{fig:box_plot}
\end{figure}
The results are presented in \cref{fig:box_plot},
showing the distribution of the accuracy of the linear approximation models
learned by LIME and R-LIME\@.
R-LIME achieved higher accuracy compared to LIME for all values of $\tau$.
This suggests that the linear classifiers learned by LIME and R-LIME
differ significantly,
and R-LIME learns a high-accuracy linear classifier adapted to the rectangular region.
Additionally, as $\tau$ increases,
the variability in the accuracy of LIME widens.
This indicates that the linear classifiers learned by LIME may not function
effectively as approximation models depending on how the region is selected.

\section{Challenges and Future Work}
 {%
  \def\imgwidth{0.47\textwidth}
  \begin{figure}[t]
    \centering
    \begin{subfigure}[t]{\imgwidth}
      \centering
      \includegraphics[width=0.7\textwidth]{visual-rlime3}
      \caption{R-LIME for balanced label distribution.}\label{fig:balanced}
    \end{subfigure}
    \begin{subfigure}[t]{\imgwidth}
      \centering
      \includegraphics[width=0.7\textwidth]{visual-rlime-imbalanced}
      \caption{R-LIME for imbalanced label distribution.
      }\label{fig:imbalanced}
    \end{subfigure}
    \caption[Behavior of R-LIME for balanced and imbalanced label distribution]{%
      Behavior of R-LIME for balanced and imbalanced label distribution.
      In case of imbalanced label distribution,
      the approximation region covers the entire input space and the
      linear approximation model always outputs the majority label.
    }
  \end{figure}
 }

@@

\section{Conclusion}
We argued that the interpreting the outputs of black-box models is crucial
for their application in sensitive domains,
and identified challenges in existing methods for local model-agnostic post-hoc
explanations of black-box classifiers.
To address these challenges,
we proposed R-LIME, a method that locally approximates the decision boundary
of a black-box classifier linearly and provides a rectangular approximation region,
which is interpretable for users due to being expressed as a conjunction of feature predicates.
We proposed an algorithm to
maximize coverage of the approximation region
as long as the accuracy of the linear approximation model exceeds a given threshold.
Comparing the outputs of LIME and R-LIME on the real-world dataset,
we demonstrated that R-LIME provides a clear application scope of the
explanation, can be evaluated by users for its reliability and generality,
and achieves higher and less variable local accuracy compared to LIME\@.
However, we discussed the instability of behavior against imbalanced label
distributions and raised questions about the theoretical validity of using
the KL-LUCB algorithm for our problem.
